---
title: "Resources and Community"
output:
  html_document:
    fig_width: 1
    toc: true
    toc_float: true
---
<!---From http://stackoverflow.com/questions/31753897/2-column-section-in-r-markdown:
Put in your css file or directly in rmarkdown--->
<style>
  .col2 {
    columns: 2 200px;         /* number of columns and width in pixels*/
    -webkit-columns: 2 200px; /* chrome, safari */
    -moz-columns: 2 200px;    /* firefox */
  }
</style>


We work and code openly and collaboratively with R, RStudio, git, and GitHub. 
We learned using the following resources, which are all available online for free, and developed and used by a welcoming community.

**Resources** 

<!--- TODO NING: format Box 1, Box 2 
- MEAM webinar: telling our story to a wider audience

--->

## Learn open data science tools 

<div class="col2">


```{r, child='box1_web.Rmd'}
```

</div>

<br>
<br>

## Strategies to learn in an intentional way

```{r, child='../opensci-manuscript/manuscript/box2.Rmd'}
```



<br>


**Build the environmental data science community**

Follow us on Twitter
<!--- 
@OHIscience
@juliesquid
@jafflerbach
@oharac
@ben_d_best
@BSHalpern
--->  
  
Note about eco-data-science

See also: 

[The importance of open data science tools in science: a list of references](http://ohi-science.org/news/importance-of-open-data-science-tools)

<!---

**Community Calls** https://github.com/ropensci/commcalls/issues

We will host community calls to share stories bridging environmental science and data science. 

- Community Call 1: Ocean Health Index Science Team: Our Path to Better Science in Less Time

--->
